Towards a standardized evaluation of imputation methodology

### 2022-03-25 Formatted/Formatted.tex
Add last version that we discussed to run latexdiff on

The problem with model-based data generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem that is being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness is subsequently induced, the observed parts of the data still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizibility of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s) and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit conditions of the data, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries is very challenging. The idea behind this data generating mechanisim is to resample repeatedly from an empirical dataset which presents certain relevant characteristics but has an unknown true data generating mechanism. Since most real-world data contains at least one missing entry, and the non-response model underlying this missingness is by definition unknown, the simulator needs to deal with missingness in \textit{some} way before empirical data can serve as comparative truth in the simulations. It may seem as an intuitive solution to only draw complete cases from the incomplete dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values in the incomplete dataset, due to the unknown non-response model. Only sampling complete cases would thus result in a simulation which fails to capture all relevant real-world conditions [which refutes the main reason for using a design-based data generating mechanism in the first place]. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset from which samples may be drawn. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations.

[Add that some simulation studies do not use a 'ground truth' at all, they just look for the method with the best predictive performance for a certain completely observed target variable. These studies apply different combinations of imputation and estimation methods on one or more benchmark datasets, and assess the RMSE of the predicted values after imputation and estimation to say something about the comparative performance of the methods. In this context, there is no talk of retrieving the 'true' but unobserved values at all. Refer to missing data chapter \citep{liu21}.]

Something else to consider is that if missing data methodology is the primary focus of the simulation, it may not be necessary to draw data repeatedly (whether it may be from a known model or empirical data). Sampling variation has always been an essential part of the evaluation of multiple imputation methodology. However, in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness, it is not necessary to take sampling variation into account \citep{vink14}. After all, we are interested only in the missing data mechanism, and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Note that the conventional pooling rules \citep[cf.][p. 76-77]{rubi87} do not apply for finite population inference. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. [An example from our own experience is that it may be difficult to get true parameter estimates of transformed data (in this case, the mean of a variable with a skew normal distribution), If we wouldn't have used a simulation set-up without sampling variance, it may have led to imprecise reference values of the estimands in our simulation. So, on the one hand, the typical set-up obfuscates the process that we're interested in (i.e., varying the missing data problem over simulation repetitions), while on the other hand, SEs may be incorrect if the sampling variance is not handled correctly in subsequent inferences.] % we hebben al sampling variance tussen de incomplete datasets, extra sampling variance op het complete data level vertroebelt het proces alleen maar. we verplaatsen liever de monte carlo naar het niveau van de missingness.

@@

In order to evaluate the effectivity of imputation methodology to recover the truth established from a chosen data generating mechanism, some form of missingness has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness patterns and missingness mechanisms. Missingness patterns concern the location of missing entries across variables in an incomplete dataset, whereas missingness mechanisms describe the relationship between missingness and the values of variables in the data \citep[][p. 8]{litt20}. Under this definition, the missingness pattern also encompasses the proportion of missingness in a given dataset.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term 'missingness proportion' restricted to the variable-by-variable interpretation. This distinction, however, is not always clear in the existing literature. Convoluting the two concepts may lead to incorrect generalizations from simulation studies, since the two are rarely equal (e.g., a proportion of 50\% incomplete cases could translate to a missingness proportion of just 25\% in bivariate data, or even 10\% in data with five variables). Moreover, the value of the actual missingness proportion could be diffusing too. Some studies use only 10\% missing data where other studies push the limits to additionally investigate the performance under at least 50\% missing data. This inconsistent display of simulation results may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of the less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

@@

Other problems with missingness patterns entail the distribution of missing values across cases (as opposed to variables). [Add: monotone/univariate vs multivariate, sporadic vs systematic, and univariately generated missing data may not have the expected multivatiate patterns.] Specific missing data patterns may inadvertently guide which imputation method prevails in performance. Some ad hoc methods are known to yield valid inferences under certain restricted conditions (e.g., list-wise deletion may outperform imputation methods if there is missingness in the outcome variable of the analysis model exclusively), while they bias inferences in any other case. Extrapolating simulation results from such a specific missing data pattern to more intricate empirical missingness patterns could lead to false conclusions.

[Missingness mechanisms: 1) definitions of mechanisms, 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated \textit{spuriously} if the correlation between variables is low, and 4) that MNAR is often ignored, while it may be the most realistic in practice. And finally something about types of M(N)AR, but what are specific problems with that? ref dance paper rianne]

% MCAR missingness: The scenario where the missing values are missing at random and the observed values are observed at random. Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation.

% Missingness type: Given the simulated data-distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single MAR mechanism.

%  the validity of assumptions become increasingly important when the missingness increases.

In order to obtain valid imputation inference, the imputation model must capture the essence of the true non-response mechanism \citep{meng94}. The model - if any - that is used to generate the missingness is usually assumed to be random (MAR) or completely random (MCAR). Missingness mechanisms that are not random (MNAR) and mechanisms that are considered non-ignorable \citep[see e.g.][]{rubi76}, are generally ignored during evaluation of imputation methodologies, except for those methods that are specifically targeted at non-ignorable applications.

With MCAR missingness mechanisms, the probability to be missing is the same for all cases. This is a necessary simulation condition for evaluating the performance of imputation procedures. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound. Sadly, the straightforward case of MCAR is often neglected from simulation studies and focus is drawn to the evaluation of MAR mechanisms only. Alternatively, some studies limit their evaluations to MCAR mechanisms only, which in our view may be far too simplistic.

With MAR missingness mechanisms, the probability to be missing is the same within groups of cases defined by the observed data only (e.g. males are less likely to disclose their weight, but gender is observed). The essence here is that the observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on gender), but these relations may be weak, or non-existent.

When MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. When MAR is induced from data without multivariate relations, the inferential implication of the missingness would mimic those of MCAR. This problem is generally overlooked in simulation studies and the procedure to generate missing data is often insufficiently described in resulting publications.

@@

A MAR missingness generated from data with low correlation would yield similar inference to the MCAR missingness, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than multiple imputation \citep[see e.g.][p. 48]{fimd}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is useless.

Although one cannot definitively verify if the missingness is random - after all, for every MNAR mechanism there is a MAR mechanism with equal fit \citep{molenberghs2008every} - it can be argued that MNAR is the more likely mechanism for real life missingness scenarios. In such cases, an indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

@@

Moreover, the performance of imputation procedures on distributional properties is often ignored in simulation studies, and even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny. Rather, one would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions.

@@

@@

Convergence of the algorithm: Most contemporary imputation techniques rely on iterative algorithms, such as the Gibbs sampler, where some algorithms are critically considered to be possibly incompatible Gibbs samplers \citep[PIGS,][]{li2012imputing}. The convergence of all iterative algorithms should always be considered and if non-convergence is suspected, the inference resulting from the imputations should not be considered.

Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data. Under anything but the MCAR assumption, this can be expected. When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed \citep[see][for an detailed overview of diagnostic evaluation for multivariate imputations]{abayomi2008diagnostics}. When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.

@@

Plausibility of the imputed values: Plausible imputations - imputations that could be real values if they had been observed - are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values.

@@

>>>>

The problem with model-based data generating mechanisms is that a method's performance on simulated data may not translate to empirical data. Real-life data hardly ever follow a given theoretical distribution, so there is no guarantee that simulation results are generalizable. Moreover, data are often generated such that the problem that is being studied is most pronounced, e.g. with consistently high correlations between groups of variables. This results in simulated data that contain such valuable information structures that, no matter what type of missingness is subsequently induced, the observed parts of the data still hold much (if not all) of the information about the missing part. Unsurprisingly, the performance of any imputation method will then be evaluated as good. Another threat to the generalizibility of model-based simulations is the use of a single model for both data generation and imputation. If data are generated following a model that is also used for imputing the data, the imputation approach will be deemed good (or better than other methods) purely due to the evaluated conditions being in favor of the problem that is studied. Other (unfair) comparative advantages in favor of a certain imputation method may occur due to characteristics of the generated data, such as the number of observations, the number of variables, the variable type(s) and the coherence between variables. In contrast to design-based studies, such characteristics are not always explicit simulation conditions, which may give a false sense of objectivity.

An obvious problem with design-based simulation is that obtaining a large dataset without missing entries can be very challenging. Most real-world data contains at least one missing entry, for which the true underlying non-response model is by default unknown. Therefore, the simulator needs to deal with missingness in \textit{some} way before incomplete empirical data can serve as comparative truth in the simulations. It may seem as an intuitive solution to only draw complete cases from the large dataset, which would indeed yield complete samples. However, there may be inherent differences between cases with and cases without any missing values, due to the unknown non-response model. Only sampling complete cases would thus result in a simulation that fails to capture all relevant real-world conditions [which refutes the main reason for using a design-based data generating mechanism in the first place]. Another way to deal with missingness in a design-based simulation is to impute the incomplete dataset once, to obtain a single completed dataset to draw samples from. Unfortunately, this practice may favor the imputation method that was used in this initial imputation step throughout any further evaluations. Nowadays, ... [Add that some simulation studies do not use a 'ground truth' at all, they just look for the method with the best predictive performance for a certain completely observed target variable. These studies apply different combinations of imputation and estimation methods on one or more benchmark datasets, and assess the RMSE of the predicted values after imputation and estimation to say something about the comparative performance of the methods. In this context, there is no talk of retrieving the 'true' but unobserved values at all. Refer to missing data chapter \citep{liu21}.]

Something else to consider is the number of samples that is drawn from the data generating mechanism. If evaulating missing data methodology is the primary focus of the simulation, it may not be necessary to draw data repeatedly. Sampling variation has always been an essential part of the evaluation of multiple imputation methodology. However, in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness, it is not necessary to take sampling variation into account \citep{vink14}. After all, we are interested only in the missing data mechanism, and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Note that the conventional pooling rules \citep[cf.][p. 76-77]{rubi87} do not apply for finite population inference. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. [An example from our own experience is that it may be difficult to get true parameter estimates of transformed data (in this case, the mean of a variable with a skew normal distribution). If we wouldn't have used a simulation set-up without sampling variance, it may have led to imprecise reference values of the estimands in our simulation. So, on the one hand, the typical set-up obfuscates the process that we're interested in (i.e., varying the missing data problem over simulation repetitions), while on the other hand, SEs may be incorrect if the sampling variance is not handled correctly in subsequent inferences.] % we hebben al sampling variance tussen de incomplete datasets, extra sampling variance op het complete data level vertroebelt het proces alleen maar. we verplaatsen liever de monte carlo naar het niveau van de missingness.

@@

In order to evaluate the effectivity of imputation methodology in recovering the established truthfrom the data generating mechanism, some form of missingness has to be induced. Often, however, reports of simulation studies remain vague about the actual missingness conditions under investigation and, even worse, some authors only report something like
\begin{quote}
\textit{We generated missing data following a missing at random missingness mechanism.}
\end{quote}
This should be considered unacceptable as claims about the validity of the imputation inference heavily depend on the simulated missingness conditions, such as missingness patterns and missingness mechanisms. Missingness patterns concern the location of missing entries across variables in an incomplete dataset, whereas missingness mechanisms describe the relationship between missingness and the values of variables in the data \citep[][p. 8]{litt20}. Under this definition, the missingness pattern also encompasses the proportion of missingness in a given dataset.

@@

Even the terminology on missingness generation can be confusing. Does a missingness proportion of 50\% mean that half of the entries in an incomplete dataset are missing, or that half of the rows have at least one missing entry? In this paper, we will henceforth refer to the latter as the proportion of incomplete cases, and keep the term 'missingness proportion' restricted to the variable-by-variable interpretation. This distinction, however, is not always clear in the literature. Convoluting the two concepts may lead to incorrect generalizations from simulation studies, since the two are rarely equal (e.g., a proportion of 50\% incomplete cases could translate to a missingness proportion of just 25\% in bivariate data, or even 10\% in data with five variables). Moreover, the value of the actual missingness proportion may be diffusing too. Some studies use only 10\% missing data [rephrase to use missingness proportion?] where other studies push the limits to additionally investigate the performance under at least 50\% missing data. This inconsistent display of simulation results may impact the objectivity of meta-evaluations over imputation methods, as one method's performance may appear to be favorable because of the less stringent simulation conditions. This ultimately may lead to statisticians recommending a less efficient method to applied researchers, thereby limiting the efficiency of the imputation approach and unnecessarily lowering the statistical power.

@@

Other problems with missingness patterns entail the distribution of missing values across cases (as opposed to the proportion over variables). [Add: monotone/univariate vs multivariate, sporadic vs systematic, and univariately generated missing data may not have the expected multivatiate patterns. For example, iterative imputation algorithms converge instantly if there is only 1 variable with missingness or monotone pattern, but require iteration in all other situations.] Specific missing data patterns may inadvertently guide which imputation method prevails in performance. Some ad hoc methods are known to yield valid inferences under certain restricted conditions (e.g., list-wise deletion may outperform imputation methods if there is missingness in the outcome variable of the analysis model exclusively), while biasing inferences in any other case. Extrapolating simulation results from such a specific missing data pattern to more intricate empirical missingness patterns could lead to false conclusions.

[Missingness mechanisms: 1) definitions of mechanisms, 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated \textit{spuriously} if the correlation between variables is low, and 4) that MNAR is often ignored, while it may be the most realistic in practice, 5) and finally something about types of M(N)AR (ref dance paper rianne).]

% MCAR missingness: The scenario where the missing values are missing at random and the observed values are observed at random. 
%  the validity of assumptions become increasingly important when the missingness increases.

\begin{table}[htb]
\begin{center}
\caption{Missingness mechanisms.}
\begin{tabular}{lll}
\hline
Mech. & Interpretation and consequences of the mechanism \\
\hline  
MCAR  & The probability to be missing is the same for all cases. In other words, the missing \\
      & values are missing at random and the observed values are observed at random. \\
MAR   & The probability to be missing is the same within groups of cases defined by the \\
      & observed data only. The essence is that observed relations in the data inform the \\
      & missingness, such that MCAR may be assumed within the observed groups. \\
MNAR  & The probability to be missing dependend on unobserved aspects of the data. The \\
      & cause of the missingness is unknown and cannot be inferred from the observed data. \\
      & This is considered non-ignorable missingness \citep[see e.g.][]{rubi76}. \\
\hline
\end{tabular}
\end{center}
\end{table}

In order to obtain valid imputation inference, the imputation model must capture the essence of the true non-response mechanism \citep{meng94}. The model--if any--that is used to generate the missingness is usually assumed to be random (MAR) or completely random (MCAR).

[MCAR is important but often ignored:] MCAR is a necessary simulation condition for evaluating the performance of imputation procedures. Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation. If an imputation method is not able to solve the problem (i.e. yield valid inference) under MCAR, the statistical properties of the procedure are not sound. Sadly, the straightforward case of MCAR is often neglected from simulation studies and focus is drawn to the evaluation of MAR mechanisms only. Alternatively, some studies limit their evaluations to MCAR mechanisms only, which in our view may be far too simplistic.

[MAR may be spurious:] With MAR missingness mechanisms, observed relations in the data are used to induce missingness during simulation (e.g. weight is made incomplete based on gender) such that the probability to be missing is the same within groups of cases defined by the observed data (e.g. males are less likely to disclose their weight, but gender is observed). The relations in the observed data may, however, be weak or non-existent. If MAR is induced based on weaker relations in the data, claims for a method's applicability to situations where the missingness is random become less valid. In the most extreme case, MAR is induced from data without multivariate relations. The inferential implications of the missingness would then mimic those of MCAR, even though the missingness is random. In fact, this would amount to one of the special cases under which complete case analysis would be more efficient than multiple imputation \citep[see e.g.][p. 48]{fimd}: the missingness does not depend on the incomplete variable. This property might be useful in practice, but considering it as a condition to evaluate performance under MAR missingness is useless.

@@

[MNAR is often ignored:] Missingness mechanisms that are not random (MNAR) and mechanisms that are considered non-ignorable \citep[see e.g.][]{rubi76}, are generally ignored during evaluation of imputation methodologies, except for those methods that are specifically targeted at non-ignorable applications. Although one cannot definitively verify if the missingness is random--after all, for every MNAR mechanism there is a MAR mechanism with equal fit \citep{molenberghs2008every}--it can be argued that MNAR is the more likely mechanism for real life missingness scenarios.

[Missingness type is often ignored (duplicated text from solutions section): Given the simulated data-distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook such results due to the focus on a single MAR mechanism.

@@

Moreover, the performance of imputation procedures on distributional properties is often ignored in simulation studies, and even though the estimates on the analysis level may be justified, some methods can yield imputations that may seem completely invalid to applied researchers. For example, one could very accurately estimate average human height by filling in negative values and values that are unrealistically large. While the obtained inference could still be valid under such imputations, the plausibility of the imputed values given the observed data should be under scrutiny.

@@

[Although often ignored, it can be argued that MNAR is the more likely mechanism for real life missingness scenarios. It would be wise to include the mechanism in your simulation study, or perform sensitivity analyses.] An indication of the validity of the obtained inference, given that the assumed missingness mechanism is suspected to be invalid, may be obtained by performing sensitivity analysis \citep[see e.g.][part 5]{molenberghs2014handbook}.

@@

Convergence of the algorithm: Most contemporary imputation techniques rely on iterative algorithms, such as the Gibbs sampler, where some algorithms are critically considered to be possibly incompatible Gibbs samplers \citep[PIGS,][]{li2012imputing}. The convergence of all iterative algorithms should always be considered and if non-convergence is suspected, the inference resulting from the imputations should not be considered. 
Distributional characteristics: In practice, the distribution of the incomplete data may differ greatly from the observed data. Under anything but the MCAR assumption, this can be expected. When evaluating imputations, the distributional shapes should be checked and diagnostic evaluations should be performed \citep[see][for an detailed overview of diagnostic evaluation for multivariate imputations]{abayomi2008diagnostics}. When anomalies are found, and if the imputation method is valid, there should be an explanation, especially in the controlled environment of a properly executed simulation study.

@@

Plausibility of the imputed values: Plausible imputations--imputations that could be real values if they had been observed--are not a necessary condition for obtaining valid inference. However, in practice, especially when the imputer and the analyst are different persons, plausibility of imputations may be a desired property. One would prefer an imputation technique to yield both valid inference and plausible imputations. It should be studied if an imputation method is prone to deliver such impractical results, and if so, under what conditions. When evaluating imputation routines, the evaluator should mention whether the routine is prone to deliver implausible values.

@@

The problem of spurious MAR is generally overlooked in simulation studies and the procedure to generate missing data is often insufficiently described in resulting publications.
